Online Social Cohesion Reflects Real-World Group Action in Syria during the Arab Spring

### 2020-06-01 manuscript.Rmd
Integrating Leah's edits pt. 1

@@

## Arab Spring
In December 2010, a Tunisian man named Mohamed Bouazizi was harassed and
unjustifiably stripped of his vendor cart by municipal officials. His
humiliation led him to set himself on fire in protest to the authoritarian
rule, ultimately leading to his death several days later. His actions served
as a catalyst for the Tunisian Revolution, and one week after his death,
Zine El Abidine Ben Ali---the longstanding authoritarian president---fled the
country. The events of Tunisia rapidly spread to surrounding countries 
also subject to unjust authoritarian rule, including Yemen, Egypt, Libya,
and Syria.

These mass movements would not have been possible without
efficient large-scale coordination, both physically on the ground and virtually 
on social media platforms. Along with Facebook, Twitter has been credited as 
essential to the Arab Spring---both for protesters (as a means to coordinate 
with one another and to communicate with the world) and for the government 
response (to monitor and track protestors to crack down on dissent) 
[@comninos2011twitter]. The events of Arab Spring have been thoroughly 
discussed within the political science literature 
[e.g., @lutterbeck2013arab; @koehler2017political], but to our knowledge,
the direct *coupling* between Syrian Twitter cohesion and international action
coordination has not been investigated.

@@

Social media platforms provide one way to enable this kind of online social
cohesion by offering distributed information-gathering and real-time information
dissemination. Given its accessibility and proficiency as an information
exchange tool, Twitter has provided individuals involved in social movements
an opportunity to complement or even replace real-world social mobilization.
For example, during the 2012 Italian protests for global economic change, the riots in
Rome were mirrored by Twitter activity both on the day of the protests and in the
days following them [@vicari2013public]. Tweeting was
determined to be more effective locally in discussing real-world, real-time
events than traditional mainstream media platforms. As more and more
individuals rely on social media to obtain local news and information, it
becomes increasingly important to understand just how these micro-blogs
reflect real-world action.

>>>>

Prior to the advent of social media, people coordinated their protest behavior using phone tree calling sheets and flyers generated on mimeograph machines, through in-person meetings for social organizations and in churches, and through word of mouth. They gathered together and hand-made protest signs that reflected their agendas. As McAdam found, people were more likely to join movements when they had a personal connection [@McAdam_1986; @McAdam_1986]. Strong ties were better predictors of involvement in high-risk protest activity. However, the era of the Internet and online social mobilization through decentralized networks has upended theories about recruitment to, and participation in, social movements. Strong interpersonal ties have yielded to weak impersonal ties that strongly predict involvement in social movements [@Threlkeld_2017]. The research methodologies for measuring social movements have similarly evolved from ethnographic interviews to big data analysis of language on social media. The cohesiveness and strength of social movements in the pre-Internet era could be operationalized by the number of people turning out to protest, and through the political events that precipitated and resulted from social gatherings en masse.

In the offline era of social mobilization, the content of placards and protest signs provided the protest narrative for onlookers near and far. For cohesive social movements, the content of placards and posters itself was cohesive. For example, the “I am a man” posters held by the men involved in the 1968 sanitation workers’ strike reflected their internal cohesion and unified, goal-oriented movement [@Windsor_2020]. Social media replicates and amplifies this phenomenon during moments of social unrest. Amid the cacophony of messages during political upheaval, such as the Arab Spring, consistent themes emerge through hashtags, retweets, and the repeated mentions of concepts such as regime, security, demonstration, and human rights. The cohesiveness of these themes, i.e., whether movement participants are “singing in one voice,” reflects the real-world events happening contemporaneously with collective action in social movements or civil conflicts.

In the present research, we investigate the relationship between linguistic cohesion and social cohesion in times of social conflict and unrest. We tackle a question related to the one posed by Zeitzoff (2017): how does social media reflect on-the-ground realities during times of social conflict [@Zeitzoff_2017]? We examine patterns of linguistic cohesion on Twitter during the Arab Spring uprising in Syria between March and June 2012. We contextualize the linguistic cohesion with social mobilization and response to the relative intensity of the conflict. Cohesion is a multi-layered concept referring to the connectedness of words, thoughts, speakers, and actions. Cohesion reflects the continuity of language and expression, and it shows the connectivity of participants in a social movement. To address the relationship between language and discourse and social movements, we adopt an interdisciplinary approach that incorporates linguistics, psychology, and political science using non-linear dynamical systems methodology to model the relationship between language and action. We find that in the early months of 2012, conflict-related events were more strongly associated with linguistic cohesion than were more neutral or peaceful events.

Disentangling the complicated relationship between mass online mobilization, and social mobilization in person, continues to be a relevant question. The political processes that transpired during and after the Arab Spring are still in the process of being understood. As Zeitzoff (2017) recognizes, social media is changing conflict, but the mechanisms and relationships are still being understood [@Zeitzoff_2017].

## Syria in context
The decades prior to the onset of the Arabs Spring in Syria were marked by social repression and fear. In 1971, Hafez al-Assad overthrew the previous leader and remained in power for thirty years, presiding during times of regional contentiousness and cooperation until his death in 2000. His son, Bashar al-Assad, who continued the regime of authoritarian leadership, assumed his office. The Assad lineage hails from a minority Islamic sect, the Alawites, in a multi-religious and multi-ethnic country comprised of Christians, Sunni and Shia Muslims, Kurds, and Palestinian refugees.

In December 2010, a Tunisian man named Mohamed Bouazizi was harassed and
unjustifiably stripped of his vendor cart by municipal officials. His
humiliation led him to set himself on fire in protest to the authoritarian
rule, ultimately leading to his death several days later. His actions served
as a catalyst for the Tunisian Revolution, and one week after his death,
Zine El Abidine Ben Ali---the longstanding authoritarian president---fled the
country. The events of Tunisia rapidly spread to surrounding countries 
also subject to unjust authoritarian rule, including Yemen, Egypt, Libya,
and Syria.

Following a wave of democratization efforts across the Middle East and North Africa, protests began modestly in Syria in March 2011. The regime responded harshly with imprisonments, detentions, censorship, and military operations as the resistance to Assad’s regime grew and spread throughout the country. The fighting escalated in earnest in March 2012, after increased unrest in the western city of Homs. International actors, including the Red Cross/Red Crescent Society, the United Nations General Assembly, the Security Council, and individual leaders throughout the world worked toward cease-fires, negotiations, and concessions, without much success. While many high-level military and diplomatic officials have defected from the regime, sufficient loyal forces have been able to execute continued repression. During this time between March and June 2012, more than eight million tweets were sent within Syria revealing the details of the conflict.

Syria is a multi-ethnic, multi-lingual, and multi-religious state. Kurdish people comprise between 10-15% of the population, Alawites between 8-15%, Christians about 10%. There are also much smaller sects, i.e. the Ismailis and Druze, but Sunni Muslims comprise the vast majority of society, and most people speak Arabic as their mother tongue. Most policy reports addressing the current conflict report on the cohesion of the ruling Alawite military elite, rather than on the cohesiveness of the opposition movement. However, the Sunni majority’s disaffection with the Syrian government and lack of political representation provide sufficient grievance and collective group identification to make them a cohesive force.

## Social Media and Social Movements

During the Arab Spring, social media provided real-time information about the events transpiring in countries across the Middle East and North Africa. Our understanding about the role of social media in social movements is still evolving. Some argue that these mass movements would not have been possible without
efficient large-scale coordination, both physically on the ground and virtually 
on social media platforms. Along with Facebook, Twitter has been credited as 
essential to the Arab Spring---both for protesters (as a means to coordinate 
with one another and to communicate with the world) and for the government 
response (to monitor and track protestors to crack down on dissent) 
[@comninos2011twitter]. The events of Arab Spring have been thoroughly 
discussed within the political science literature 
[e.g., @lutterbeck2013arab; @koehler2017political], but to our knowledge,
the direct *coupling* between Syrian Twitter cohesion and international action
coordination has not been investigated. Others provide skepticism about the effect of social media on collective action [@Seay_2014].

In Foreign Affairs, Anderson writes that, “The important story about the 2011 Arab revolts in Tunisia, Egypt, and Libya is not how the globalization of the norms of civic engagement shaped the protesters' aspirations. Nor is it about how activists used technology to share ideas and tactics. Instead, the critical issue is how and why these ambitions and techniques resonated in their various local contexts [@Anderson_2011].” Malcolm Gladwell, in highlighting the notions of strong and weak ties, @Gladwell_2010 and @McAdam_1986, similarly expressed hesitation about the role of online organizing, arguing that real-life strong ties outweigh online weak ties and that participation in online activism only requires scant sacrifices whereas recruitment to higher-risk activities is done through in-person contact, namely strong ties [@Gladwell_2010; [@McAdam_1986]. Interestingly, these views may actually help to make the case for bridging the gap between social media and social mobilization, by suggesting the strength of social ties in an online environment is inextricably linked to the context and ambitions of the actors. Granovetter notes that while strong ties tend to bind smaller groups, weak ties tend to bind groups across distances [@Granovetter_1973].

Social media generates critical information about events, such as whether or not a demonstration will take place and what transpires when it does, how many people participate, and how the state, the military, and the security apparatus respond. Similarly, these media platforms alert us to more global shifts within a conflict, such as when fighting moves to a new city or region, the number of casualties and wounded, and the types of weaponry used by parties to the conflict [@Zeitzoff_2017]. Social media is not entirely event-driven, however. In monitoring social media channels, government forces may respond to the perception that citizens are gaining strength and sympathy. Governments have been shown to be quite tolerant of dissent so long as it does not metamorphosize into collective action [@King_Pan_Roberts_2013]. Additionally, the insulated nature of online social platforms allows valuable social dynamics to be detailed at unprecedented resolution and scale. As such, the digital traces left by individuals and collectively as groups are regarded as a valuable source of information that can offer powerful insights into states of social disequilibrium, up-risings, and escalations into civil wars.

Citizens communicate with each other with increased transparency due to the ubiquitous availability of social media such as low-cost and egalitarian platforms like Twitter. Citizens are able to access these tools via mobile phones and computers with Internet connectivity, providing a stream of real-time shared information. This facilitates studying the degree of cohesion between speakers, and the alignment between that cohesion and their collective political behavior. Citizens can tweet the location of events like protests and military incursions, the degree of state repression, and the number of casualties. In kind, the government, interested in remaining in power, monitors social media platforms to ascertain emerging threats, such as where new groups are forming and where existing groups are strengthening, where gatherings are held, and who the idea entrepreneurs are. An example of a leader manifesting his interest in such public conversation is given in a speech of June 3, 2012 by Bashar al-Assad:

“I know that I have been away from the media for a long time, and I have missed having direct contact with the citizens, but I have always been following up with the daily occurrences and gathering the information so that my speech can be built on what is said by the street.”

With few exceptions, Assad chose to keep social media platforms up and running. Gohdes (2015) notes that regimes can attempt to thwart the growth of social protest online and in the streets by shutting down the Internet [@Gohdes_2015]. During these times, however, she finds that military repression increases. However, government information about the citizens decreases, and as such, many regimes have elected to maintain Internet connectivity as a means of monitoring and surveilling their citizens.

## Sociolinguistic cohesion: Connectedness of speakers, events, and ideas
We tie the constructs of lexical, interlocutor, and social/physical cohesion together in a theoretical framework called sociolinguistic cohesion, as shown in Figure X. We can measure the changes in cohesion preceding and in response to external stimuli by evaluating the messages and the context in which they occur. Several mechanisms are in motion here. Previous values of linguistic cohesion should influence future values of linguistic cohesion. They should also influence future actions in the real world space. Previous events influence future events, and also the future cohesiveness of the opposition.

Sociolinguistic cohesion is not a static concept. Just as social movements experience increased participation as well as setbacks – indicating dynamic relationships between participants, and participants and the sociopolitical environment – sociolinguistic cohesion is also dynamic, reflecting the changing states of the political sphere. Other scholars have noted the diffusion of messages whereby they reach a critical mass, whereby “behaviors, beliefs, and preferences” spread within a population [@Barash_Cameron_Macy_2012].  Cohesion changes over time in response to external events and the internal dynamics of the social movement. Some may wonder whether social cohesion can diffuse among distant actors who do not interact face-to-face; Centola and Macy show that Internet weak ties extend over long distances and as such have spatially significant abilities to influence actors [@Centola_Macy_2007]. Fink et al. also explored diffusion of hashtags in viral contagion in Nigeria, a phenomenon similar to what we identify as sociolinguistic cohesion [@Fink_Schmidt_Barash_Cameron_Macy_2016]. The prevalence and rate of particular hashtags contribute to the overall sociolinguistic cohesion. he more that people reuse hashtags or common phrases, the greater the sociolinguistic cohesion when this phenomenon occurs during socially significant times like protests [@Adamic_others_2015]. Exogenous events, like intervention by another country or international organization like the United Nations, can affect sociolinguistic cohesion. Endogenous events, like the recruitment of new members to the rebel or resistance movement, can also affect the level of cohesion.

Exploring how social media influences contentious behavior, Tilly notes that social media lowers coordination and transaction costs for participants, and that older versions of organizing social movements have become less effective, yielding to the new frontier of online organizing [@McAdam_Tarrow_Tilly_2003]. Koopmans extends this further, writing that, “Authorities react to social movement activities if and as they are depicted in the mass media, and conversely movement activists become aware of political opportunities and constraints through the reactions (or non-reactions) that their actions provoke in the public sphere” [@Koopmans_2004].

Lexical cohesion, then, should be a useful indicator of social cohesion in political crisis. Lexical cohesion happens at the word level. Interlocutor cohesion happens between speakers. Political cohesion happens at the social level during moments of political crisis, like civil wars and social movements. Sociolinguistic cohesion is comprised of lexical, interlocutor, and socio-political events, where the words and their context reveal the disposition of socially connected linguistic communities. Sociolinguistic cohesion combines the elements of lexical cohesion in that it captures the degree of similarity between what is said, and how it is said. It accounts for the cohesion between multiple speakers in an environment whose language evolves to become more aligned and similar. Sociolinguistic cohesion also addresses how groups involved in political rebellion must have internal cohesion to overcome the collective action problem to mount sustained resistance to state authority.

@@

@@

This kind of massive online social cohesion can lead to real-world
social manifestations. Social media platforms provide one way to enable this kind of online social
cohesion by offering distributed information-gathering and real-time information
dissemination. Given its accessibility and proficiency as an information
exchange tool, Twitter has provided individuals involved in social movements
an opportunity to complement or even replace real-world social mobilization.
For example, during the 2012 Italian protests for global economic change, the riots in
Rome were mirrored by Twitter activity both on the day of the protests and in the
days following them [@vicari2013public]. Tweeting was
determined to be more effective locally in discussing real-world, real-time
events than traditional mainstream media platforms. As more and more
individuals rely on social media to obtain local news and information, it
becomes increasingly important to understand just how these micro-blogs
reflect real-world action.
